---
title: "Intro to Formal Political Analysis:\\newline Strategic Games and Nash Equilibria in Pure Strategies, Part One"
author: "JBrandon Duck-Mayr"
date: "`r quack::american_date_format(Sys.Date())`"
urlcolor: Blue
header-includes:
    - \newcommand{\setsep}{\setlength{\itemsep}{3pt}}
    - \newcommand{\setskip}{\setlength{\parskip}{3pt}}
    - \renewcommand{\tightlist}{\setsep\setskip}
    - \setlength{\abovedisplayskip}{-6pt}
output:
    quack::presentation:
        toc: false
        incremental: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

## What is a "strategic game"?

. . .

::: {.block}
### Osborne, Definition 13.1
A strategic game consists of

\vspace*{3pt}

. . .

- a set of **players**
- for each player, a set of **actions**
- for each player, **preferences** over the set of action profiles
:::

. . .

&nbsp;

This is also called a "normal-form game"

## Example: The Prisoner's Dilemma

. . .

- Two suspects in a major crime are questioned separately
- The cops already have enough evidence to convict them of a minor crime but need a confession to convict on the major crime
- So if neither talks, they both get convicted of the minor crime but neither gets convicted of the major crime
- If only one of them talks, that prisoner will get a plea deal and serve no time while the other will be convicted of the major crime
- If both talk, they are both convicted of the major crime, but get a slightly reduced sentence for confessing

## The Prisoner's Dilemma as a normal form game

. . .

- Players: The two suspects
- Actions: Each player can stay quiet ($C$) or talk ($D$)
- Preferences:
  + From best to worst, Player 1 prefers:
    - $(D, C)$; 1 goes free and 2 is convicted of the major crime
    - $(C, C)$; both are only convicted of the minor crime only
    - $(D, D)$; both are convicted of the felony with a reduced sentence
    - $(C, D)$; 1 is convicted of the major crime and pays full freight
  + Player 2's ordering is $(C, D)$, $(C, C)$, $(D, D)$, and $(D, C)$.

. . .

&nbsp;

A utility function that represents Player 1's preferences is $$u_1(D, C) = 3,\ u_1(C, C) = 2,\ u_1(D, D) = 1,\ u_1(C, D) = 0$$

## Matrix representation of The Prisoner's Dilemma

. . .

```{r prisoners-dilemma-matrix}
library(kableExtra)
game = data.frame(
    C = c("$2, 2$", "$3, 0$"),
    D = c("$0, 3$", "$1, 1$"),
    row.names = c("C", "D")
)
kable(game, escape = FALSE, align = "c") %>% kable_styling(position = "center")
```

## Matrix representation of The Prisoner's Dilemma

I produced that for these slides using R Markdown with the code:

````{verbatim, echo = TRUE}
```{r prisoners-dilemma-matrix}
library(kableExtra)
game = data.frame(
    C = c("$2, 2$", "$3, 0$"),
    D = c("$0, 3$", "$1, 1$"),
    row.names = c("C", "D")
)
kable(game, escape = FALSE, align = "c") %>%
    kable_styling(position = "center")
```
````

## Matrix representation of The Prisoner's Dilemma

You could produce a similar payoff matrix in LaTeX with the code:

````{verbatim, lang = "latex", echo = TRUE}
\begin{game}{2}{2}
    & $C$       & $D$       \\
$C$ & $2, 2$    & $0, 3$    \\
$D$ & $3, 0$    & $1, 1$
\end{game}
````

(this requires you add `\usepackage{sgame}` to your preamble)



## Choosing an action in the Prisoner's Dilemma

::: nonincremental

- Players: The two suspects
- Actions: Each player can stay quiet ($C$) or talk ($D$)
- Preferences:
  + From best to worst, Player 1 prefers:
    - $(D, C)$; 1 goes free and 2 is convicted of the major crime
    - $(C, C)$; both are only convicted of the minor crime only
    - $(D, D)$; both are convicted of the felony with a reduced sentence
    - $(C, D)$; 1 is convicted of the major crime and pays full freight
  + Player 2's ordering is $(C, D)$, $(C, C)$, $(D, D)$, and $(D, C)$.

:::

&nbsp;

A utility function that represents Player 1's preferences is $$u_1(D, C) = 3,\ u_1(C, C) = 2,\ u_1(D, D) = 1,\ u_1(C, D) = 0$$

## Nash Equilibrium

. . .

::: {.block}
### Osborne, Definition 23.1
The action profile $a^\ast$ in a strategic game... is a **Nash equilibrium** if, for every player $i$ and every action $a_i$ of player $i$, $a^\ast$ is at least as good according to player $i$'s preferences as the action profile $(a_i, a_{-i}^\ast)$ in which player $i$ chooses $a_i$ while every other player $j$ chooses $a_j^\ast$. Equivalently, for every player $i$,

$$u_i\left(a^\ast\right) \geq u_i\left(a_i, a_{-i}^\ast\right) \text{ for every action } a_i \text{ of player } i,$$

where $u_i$ is a payoff function that represents player $i$'s preferences.
:::

. . .

In other words, a Nash equilibrium is an action profile where no player can gain by **unilateral** deviation

## Nash equilibrium in The Prisoner's Dilemma

\begin{table}
\begin{tabular}{lcc}
& C & D \\
C & 2, 2 & 0, \textbf<4->{3} \\
D & \textbf<2->{3}, 0 & \textbf<3->{1}, \textbf<5->{1}
\end{tabular}
\end{table}

\phantom{
\parbox{\linewidth}{
\vspace*{\baselineskip}
\begin{itemize}
\tightlist
\item $(D, D)$ is the only profile where neither can gain by deviating
\item It is therefore the unique Nash equilibrium
\item Notice they'd both be better off in the action profile $(C, C)$
\item But Nash equilibrium is about unilateral deviation
\end{itemize}
}
}

## Nash equilibrium in The Prisoner's Dilemma

\begin{table}
\begin{tabular}{lcc}
& C & D \\
C & \textcolor<3->{BurntOrange}{\textbf<3->{2, 2}} & 0, \textbf{3} \\
D & \textbf{3}, 0 & \textbf{1}, \textbf{1}
\end{tabular}
\end{table}

&nbsp;

- $(D, D)$ is the only profile where neither can gain by deviating
- It is therefore the unique Nash equilibrium
- Notice they'd both be better off in the action profile $(C, C)$
- But Nash equilibrium is about *unilateral* deviation

## Strict domination

. . .

::: {.block}
### Osborne, Definition 45.1
In a strategic game... player $i$'s action $a_i''$ **strictly dominates** her action $a_i'$ if

$$ u_i\left(a_i'', a_{-i}\right) > u_i\left(a_i', a_{-i}\right), $$

for every list $a_{-i}$ of the other players' actions,
where $u_i$ is a payoff function that represents player $i$'s preferences.
We say that the action $a_i'$ is **strictly dominated**.
:::

## Weak domination

::: {.block}
### Osborne, Definition 46.1
In a strategic game with ordinal preferences, player $i$'s action $a_i$ **weakly dominates** her action $a_i'$ if

$$
u_i \left(a_i , a_{-i} \right) \geq u_i \left(a_i' , a_{-i} \right)
$$

for every list $a_{-i}$ of the other players' actions and

$$
u_i \left(a_i , a_{-i} \right) > u_i \left(a_i' , a_{-i} \right)
$$

for some list $a_{-i}$ of the other players' actions...
:::

## Domination in The Prisoner's Dilemma

\begin{table}
\begin{tabular}{lcc}
& C & D \\
C & 2, 2 & 0, \textbf<4->{3} \\
D & \textbf<2->{3}, 0 & \textbf<3->{1}, \textbf<5->{1}
\end{tabular}
\end{table}

\onslide<6>{
\vspace*{\baselineskip}
Whether Player 1 plays $C$ or $D$, Player 2 strictly prefers to play $D$.
Therefore, for Player 2, $D$ \textbf{strictly dominates} $C$.
}

## Another example of domination

\begin{table}
\begin{tabular}{lcc}
  & L & R \\
T & \textcolor<3>{Blue}{\textbf<2-3>{1}} & \textcolor<3>{Blue}{\textbf<2-3>{0}} \\
M & \textcolor<3>{BurntOrange}{\textcolor<5>{Blue}{\textbf<2-5>{2}}} & \textcolor<3>{BurntOrange}{\textcolor<5>{Blue}{\textbf<2-5>{1}}} \\
B & \textcolor<5>{BurntOrange}{\textbf<4-5>{3}} & \textcolor<5>{BurntOrange}{\textbf<4-5>{2}}
\end{tabular}
\end{table}

\onslide<6>{
\vspace*{\baselineskip}
For Player 1, $M$ \textbf{strictly dominates} $T$, and $B$ \textbf{strictly dominates} $M$.
}

## Example: Bach or Stravinski?

Two people want to go out on Friday.
There's a concert playing music by Bach and a concert playing music by Stravinski.
One person prefers Bach while the other prefers Stravinski.
For both of them, the worst option is going to a concert alone.

## Example: Bach or Stravinski?

- Players: The two people
- Actions: Each player can go to the Bach concert ($B$) or the Stravinski concert ($S$)
- Preferences:
  + From best to worst, Player 1 prefers $(B, B), (S, S), (B, S), (S, B)$
  + From best to worst, Player 2 prefers $(S, S), (B, B), (S, B), (B, S)$
  + Assuming they don't care about which concert they go to if alone, a utility function that represents Player 1's preferences is $$u_1(B, B) = 2,\ u_1(S, S) = 1,\ u_1(B, S) = u_1(S, B) = 0$$

## Example: Bach or Stravinski?

\begin{table}
\begin{tabular}{lcc}
& B & S \\
B & \textbf<2->{2}, \textbf<4->{1} & 0, 0 \\
S & 0, 0 & \textbf<3->{1}, \textbf<5->{2}
\end{tabular}
\end{table}

\onslide<6->{
In both the $(B, B)$ and $(S, S)$ profiles, neither player can gain from unilateral deviation.

Therefore, both action profiles are a Nash equilibrium.
}

\onslide<7>{
Does any of Player 1's actions weakly or strictly dominate any of their other actions?
How about for Player 2?
}

## Example: Stag Hunt

::: {.block}
### Osborne, page 20 in the 9th printing
A sentence in *Discourse on the origin and foundations of inequality among men* (1755) by the philosopher Jean-Jacques Rousseau discusses a group of hunters who wish to catch a stag... They will succeed if they all remain sufficiently attentive, but each is tempted to desert her post and catch a hare.
:::

## Example: Stag Hunt

- Players: The hunters
- Actions: Each can remain attentive to the stag ($S$) or desert and catch a hare ($H$)
- Preferences: Each best prefers a profile where every player chooses $S$, next prefers any profile where they personally choose $H$, and lastly prefer any profile where they choose $S$ and at least one other player chooses $H$

## Example: Stag Hunt

A matrix representation of the two-person stag hunt is

\begin{table}
\begin{tabular}{lcc}
& S & H \\
S & \textbf<2->{2}, \textbf<4->{2} & 0, 1 \\
H & 1, 0 & \textbf<3->{1}, \textbf<5->{1}
\end{tabular}
\end{table}

\onslide<6->{
In both the $(S, S)$ and $(H, H)$ profiles, neither player can gain from unilateral deviation.

Therefore, both action profiles are a Nash equilibrium.
}

\onslide<7>{
Does any of Player 1's actions weakly or strictly dominate any of their other actions?
How about for Player 2?
}

## Example: Matching Pennies

Two people simultaneously show each other a side of a penny.
If they show each other the same side, Player 2 pays Player 1 $1, while if they show each other different sides, Player 1 pays Player 2 $1.

## Example: Matching Pennies

- Players: The two people
- Actions: Each player can show heads ($H$) or show tails ($T$)
- Preferences:
  + From best to worst, Player 1 prefers $(H, H)$ or $(T, T)$, then $(H, T)$ or $(T, H)$
  + From best to worst, Player 2 prefers $(H, T)$ or $(T, H)$, then $(H, H)$ or $(T, T)$
  + We can represent their preferences by how much money they gain or lose; $$ u_1(H, H) = u_1(T, T) = 1,\ u_1(H, T) = u_1(T, H) = -1 $$

## Example: Matching Pennies

\begin{table}
\begin{tabular}{lcc}
& H & T \\
H & \textbf<2->{1}, -1 & -1, \textbf<4->{1} \\
T & -1, \textbf<5->{1} & \textbf<3->{1}, -1
\end{tabular}
\end{table}

\onslide<6->{
There is no pure strategy Nash equilibrium in the Matching Pennies game.
}

\onslide<7>{
Does any of Player 1's actions weakly or strictly dominate any of their other actions?
How about for Player 2?
}

## Example: Voting

Consider the following scenario:

- Two candidates, $A$ and $B$, run for office
- There are $N$ voters, with $N$ odd; some of them (a majority) prefer candidate $A$ to candidate $B$, and the rest prefer $B$ to $A$.
- There is an election with mandatory voting; the candidate with the most votes wins

## Example: Voting

- Players: The $N$ voters
- Actions: Each can vote for $A$ or vote for $B$
- Preferences:
  + All players are indifferent between action profiles where $A$ wins
  + All players are indifferent between action profiles where $B$ wins
  + Players who prefer candidate $A$ prefer any action profile where $A$ wins to any action profile where $B$ wins, and vice versa for voters who prefer candidate $B$

## Example: Voting

Questions:

- Do the players have an action that dominates their other action?
- Is there any Nash equilibrium where candidate $A$ wins the election?
- Is there any Nash equilibrium where candidate $B$ wins the election?

## Up next

- For next time
  + Review what we've gone over
  + Additionally read Sections 2.8.1 to 2.8.3 in Osborne (pages 35--42 in the 9th printing)
- Don't forget to turn in Problem Set 1 before the next class!
- If you need help, my office hours are 9 a.m.--12 p.m. on Tuesdays
