---
title: "Intro to Formal Political Analysis:\\newline Strategic Games and Nash Equilibria in Pure Strategies, Part Two"
author: "JBrandon Duck-Mayr"
date: "`r quack::american_date_format(Sys.Date())`"
urlcolor: Blue
header-includes:
    - \newcommand{\setsep}{\setlength{\itemsep}{3pt}}
    - \newcommand{\setskip}{\setlength{\parskip}{3pt}}
    - \renewcommand{\tightlist}{\setsep\setskip}
    - \setlength{\abovedisplayskip}{-6pt}
    - \newcommand{\pd}[1]{\ensuremath{\frac{\partial}{\partial{#1}}}}
    - \newcommand{\gray}{\color{white}}
output:
    quack::presentation:
        toc: false
        incremental: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
quack::attach_common_packages()
```

## Recap

. . .

- A strategic or normal-form game consists of
  + Players
  + For each player, a set of actions
  + For each player, preferences over the set of action profiles
- Nash equilibrium: quality of an action profile
  + Action profile is NE if no player can gain from unilateral deviation
- Domination: comparison between two actions of one player
  + Strict domination: Action A strictly dominates action B if the player strictly prefers all action profiles where they play A to all action profiles where they play B
  + Weak domination: A weakly dominates B if we change "strictly prefers" to "weakly prefers" **and** require the preference is strict in at least one action profile

## Dominance Solvable Games

. . .

- A game is **dominance solvable** if iterated elimination of dominated actions leaves only one actions left for each player
- If you are eliminating only strictly dominated strategies, the remaining action profile is the unique Nash equilibrium to the game (IESDS / IDSDS)
  + Order of elimination does **not** matter
- If you are also eliminating weakly dominated strategies, the remaining action profile will be a Nash equilibrium, but you may have eliminated other Nash equilibria (IEWDS / IDWDS)
  + Order of elimination **does** matter

## Dominance Solvable Games: Examples

\begin{figure}
\centering
The Prisoner's Dilemma
\end{figure}
\begin{table}
\begin{tabular}{lcc}
& \textcolor<5>{LightGray}{C} & \textbf<5>{D} \\
\textcolor<3->{LightGray}{C} &
\textcolor<2>{Blue}{\textcolor<3->{LightGray}{\textbf<2>{2}}}\textcolor<3->{LightGray}{,} \textcolor<3->{LightGray}{2} &
\textcolor<2>{Blue}{\textcolor<3->{LightGray}{\textbf<2>{0}}}\textcolor<3->{LightGray}{,} \textcolor<3->{LightGray}{3} \\
\textbf<5>{D} &
\textcolor<2>{DarkGreen}{\textcolor<5>{LightGray}{\textbf<2>{3}}}\textcolor<5>{LightGray}{,} \textcolor<4>{Blue}{\textcolor<5>{LightGray}{\textbf<4>{0}}} &
\textcolor<2>{DarkGreen}{\textbf<2, 5>{1}}\textbf<5>{, }\textcolor<4>{DarkGreen}{\textbf<4, 5>{1}}
\end{tabular}
\end{table}

## Dominance Solvable Games: Examples

\begin{figure}
\centering
Bach or Stravinski
\end{figure}
\begin{table}
\begin{tabular}{lcc}
& B & S \\
B & 2, 1 & 0, 0 \\
S & 0, 0 & 1, 2
\end{tabular}
\end{table}

## Dominance Solvable Games: Examples

\begin{table}
\begin{tabular}{lccc}
& L & C & R \\
T & 0, 0 & 1, 0 & 1, 1 \\
M & 1, 0 & 1, 1 & 3, 2 \\
B & 1, 0 & 2, 1 & 2, 2
\end{tabular}
\end{table}

## Dominance Solvable Games: Examples

\begin{table}
\begin{tabular}{lccc}
& L & C & R \\
T & 0, 0 & 1, 0 & 1, 1 \\
M & 1, 1 & 1, 1 & 3, 2 \\
B & 1, 1 & 2, 1 & 2, 2
\end{tabular}
\end{table}

## More Voting: The Hotelling-Downs Model

. . .

Consider the following model of elections:

- Two candidates seek election & *only* care about winning office
- (They prefer winning to tying and prefer tying to losing)
- Suppose there is a unidimensional policy space
- The candidates run for office on a platform (a point)
- There are a continuum of voters each with an ideal point
- They vote for the candidate closest to their ideal point
- (If multiple platforms are equidistant from a point, the candidates evenly split the relevant voters)

## More Voting: The Hotelling-Downs Model

Questions:

- Who are the players?
- Are you already familiar with the game's solution?
- How would you find the solution to this game?

## Even More Voting! The Calvert-Wittman Model

- What if the candidates cared about policy rather than winning?
- (These are sometimes called "citizen-candidates")
- Does this change the Nash equilibrium? (Focus on the case where the two candidates' ideal points are on opposite sides of the median)

## Even More Voting!!!

We could think about lots more extensions or tweaks:

. . .

- What about asymmetric voter preferences?
- What about "valence"?
- What about stochastic voting?
- What about multiple dimensions?
- Etc, etc, etc

## Best Response Functions

The \textbf{best response function} of player $i$ is $$
B_i\left(a_{-i}\right) = \{ a_i \in A_i : u_i\left(a_i, a_{-i}\right) \geq u_i\left(a_i', a_{-i}\right) \forall\ a_i' \in A_i \}$$

\invisible{
\begin{block}{Osborne Proposition 36.1}
The action profile $a^\ast$ is a Nash equilibrium of a strategic game... if and only if every player's action is a best response to the other player's actions:

$$
a_i^\ast \text{ is in } B_i\left(a_{-i}^\ast\right) \text{ for every player } i.
$$
\end{block}
}

## Best Response Functions

The \textbf{best response function} of player $i$ is $$
\mathbf{B_i\left(a_{-i}\right)} = \{ a_i \in A_i : u_i\left(a_i, a_{-i}\right) \geq u_i\left(a_i', a_{-i}\right) \forall\ a_i' \in A_i \}$$

\invisible{
\begin{block}{Osborne Proposition 36.1}
The action profile $a^\ast$ is a Nash equilibrium of a strategic game... if and only if every player's action is a best response to the other player's actions:

$$
a_i^\ast \text{ is in } B_i\left(a_{-i}^\ast\right) \text{ for every player } i.
$$
\end{block}
}

## Best Response Functions

The \textbf{best response function} of player $i$ is $$
B_i\left(a_{-i}\right) = \textcolor{BurntOrange}{\Bigl\{} a_i \in A_i : u_i\left(a_i, a_{-i}\right) \geq u_i\left(a_i', a_{-i}\right) \forall\ a_i' \in A_i \textcolor{BurntOrange}{\Bigr\}}$$

\invisible{
\begin{block}{Osborne Proposition 36.1}
The action profile $a^\ast$ is a Nash equilibrium of a strategic game... if and only if every player's action is a best response to the other player's actions:

$$
a_i^\ast \text{ is in } B_i\left(a_{-i}^\ast\right) \text{ for every player } i.
$$
\end{block}
}

## Best Response Functions

The \textbf{best response function} of player $i$ is $$
B_i\left(a_{-i}\right) = \{ a_i \in A_i : u_i\left(a_i, a_{-i}\right) \geq u_i\left(a_i', a_{-i}\right) \forall\ a_i' \in A_i \}$$

::: {.block}
### Osborne Proposition 36.1
The action profile $a^\ast$ is a Nash equilibrium of a strategic game... if and only if every player's action is a best response to the other player's actions:

$$
a_i^\ast \text{ is in } B_i\left(a_{-i}^\ast\right) \text{ for every player } i.
$$
:::

## Best Response Functions: Voting example

- Let's revisit the Hotelling-Downs model of elections
<!-- - We can characterize each candidate's best response function by -->
<!--   + Determining the exhaustive list of cases for the other player's move -->
<!--   + Holding the other player's move constant in one of those cases, -->
<!--     determine the best response to that move -->
- Candidate 2's platform $x_2$ could be
  + $x_2 < m$
  + $x_2 = m$
  + $x_2 > m$

\onslide<6>{
$$
B_1\left(x_2\right) =
\begin{cases}
\left\{ x_1 : x_2 < x_1 < 2m - x_2 \right\} & \text{ if } x_2 < m \\
\left\{ m \right\} & \text{ if } x_2 = m \\
\left\{ x_1 : 2m - x_2 < x_1 < x_2 \right\} & \text{ if } x_2 < m
\end{cases}
$$
}

## Best Response Functions: Voting example

```{r voting-br1, out.width = "70%", fig.width = 3, fig.height = 3, fig.align = "center"}
dots = data.frame(x1 = c(0, 0, 1, 1), x2 = c(0, 1, 0, 1))
segs = data.frame(x1 = c(0, 0), x2 = c(0, 1), x11 = c(1, 1), x21 = c(0, 1))
shp1 = data.frame(x1 = c(0, 0.5, 1), x2 = c(1, 0.5, 1))
shp2 = data.frame(x1 = c(0, 0.5, 1), x2 = c(0, 0.5, 0))
blue = "dodgerblue4"
ggplot(data = dots, mapping = aes(x = x1, y = x2)) +
    geom_polygon(data = shp1, fill = blue, alpha = 1/2) +
    geom_polygon(data = shp2, fill = blue, alpha = 1/2) +
    geom_text(
        aes(x = 0.5, y = 0.75),
        label = "B[1](x[2])", parse = TRUE,
        color = blue, fontface = "bold"
    ) +
    geom_point(aes(x = 0.5, y = 0.5), shape = 19, color = blue, size = 2) +
    geom_segment(
        mapping = aes(xend = x11, yend = x21), data = segs,
        color = blue, linewidth = 1
    ) +
    geom_point(shape = 21, color = blue, fill = "white", size = 2) +
    scale_x_continuous(breaks = 0.5, labels = "m") +
    scale_y_continuous(breaks = 0.5, labels = "m") +
    theme_classic()
```

## Best Response Functions: Voting example

```{r voting-br2, out.width = "70%", fig.width = 3, fig.height = 3, fig.align = "center"}
seg2 = data.frame(x2 = c(0, 0), x1 = c(0, 1), x21 = c(1, 1), x11 = c(0, 1))
shp3 = data.frame(x2 = c(0, 0.5, 1), x1 = c(1, 0.5, 1))
shp4 = data.frame(x2 = c(0, 0.5, 1), x1 = c(0, 0.5, 0))
red  = "firebrick3"
ggplot(data = dots, mapping = aes(x = x1, y = x2)) +
    geom_polygon(data = shp3, fill = red,  alpha = 1/2) +
    geom_polygon(data = shp4, fill = red,  alpha = 1/2) +
    geom_text(
        aes(y = 0.5, x = 0.75),
        label = "B[2](x[1])", parse = TRUE,
        color = red, fontface = "bold"
    ) +
    geom_point(aes(x = 0.5, y = 0.5), shape = 19, color = red, size = 2) +
    geom_segment(
        mapping = aes(xend = x11, yend = x21), data = seg2,
        color = red, linewidth = 1
    ) +
    geom_point(shape = 21, color = red, fill = "white", size = 2) +
    scale_x_continuous(breaks = 0.5, labels = "m") +
    scale_y_continuous(breaks = 0.5, labels = "m") +
    theme_classic()
```

## Best Response Functions: Voting example

```{r voting-both-br, out.width = "70%", fig.width = 3, fig.height = 3, fig.align = "center"}
purp = "#6e3a58"
ggplot(data = dots, mapping = aes(x = x1, y = x2)) +
    geom_polygon(data = shp1, fill = blue, alpha = 1/2) +
    geom_polygon(data = shp2, fill = blue, alpha = 1/2) +
    geom_polygon(data = shp3, fill = red,  alpha = 1/2) +
    geom_polygon(data = shp4, fill = red,  alpha = 1/2) +
    geom_text(
        aes(x = 0.5, y = 0.75),
        label = "B[1](x[2])", parse = TRUE,
        color = blue, fontface = "bold"
    ) +
    geom_text(
        aes(y = 0.5, x = 0.75),
        label = "B[2](x[1])", parse = TRUE,
        color = red, fontface = "bold"
    ) +
    geom_point(aes(x = 0.5, y = 0.5), shape = 19, color = purp, size = 2) +
    geom_segment(
        mapping = aes(xend = x11, yend = x21), data = segs,
        color = blue, linewidth = 1
    ) +
    geom_segment(
        mapping = aes(xend = x11, yend = x21), data = seg2,
        color = red, linewidth = 1
    ) +
    geom_point(shape = 21, color = purp, fill = "white", size = 2) +
    scale_x_continuous(breaks = 0.5, labels = "m") +
    scale_y_continuous(breaks = 0.5, labels = "m") +
    theme_classic()
```

## Best Response Functions: Public good example

- Now let's consider a public good problem (Osborne Ex. 44.1)
- Two people with wealth $w$ each choose a contribution level $c_i$ to the public good
- Suppose they value the public good itself ($c_1 + c_2$) as well as their own private consumption ($w - c_i$) and value their private consumption more as the amount of the public good increases

\onslide<4>{
\begin{align*}
u_i\left(c_1, c_2\right)
& = c_1 + c_2 + w - c_i + (w - c_i) (c_1 + c_2) \\
& = w + c_j + (w - c_i) (c_1 + c_2)
\end{align*}
}

## Best Response Functions: Public good example

::: {.nonincremental}

- Now let's consider a public good problem (Osborne Ex. 44.1)
- Two people with wealth $w$ each choose a contribution level $c_i$ to the public good
- Suppose they value the public good itself ($c_1 + c_2$) as well as their own private consumption ($w - c_i$) and value their private consumption more as the amount of the public good increases: $$ u_i\left(c_1, c_2\right) = w + c_j + (w - c_i) (c_1 + c_2) $$
- How would you find the players' best response functions?

:::

## Best Response Functions: Public good example

\begin{align*}
u_i\left(c_i, c_j\right) & = w + c_j + (w - c_i) (c_i + c_j) \\
& = w + c_j + w c_i + w c_j - c_i c_j - c_i^2 \\
\gray\pd{c_i} u_i
&\gray = \pd{c_i} w c_i - \pd{c_i} c_i c_j - \pd{c_i} c_i^2 \\
&\gray = w - c_j - 2 c_i \\
\gray 0 &\gray \equiv w - c_j - 2 c_i \\
\gray 2 c_i &\gray = w - c_j \\
\gray c_i^\ast &\gray = \frac{1}{2} \left( w - c_j \right)
\end{align*}

## Best Response Functions: Public good example

\begin{align*}
u_i\left(c_i, c_j\right) & = w + c_j + (w - c_i) (c_i + c_j) \\
& = w + c_j + w c_i + w c_j - c_i c_j - c_i^2 \\
\pd{c_i} u_i & = \pd{c_i} w c_i - \pd{c_i} c_i c_j - \pd{c_i} c_i^2 \\
& = w - c_j - 2 c_i \\
\gray 0 & \gray\equiv w - c_j - 2 c_i \\
\gray 2 c_i & \gray = w - c_j \\
\gray c_i^\ast & \gray = \frac{1}{2} \left( w - c_j \right)
\end{align*}

## Best Response Functions: Public good example

\begin{align*}
u_i\left(c_i, c_j\right) & = w + c_j + (w - c_i) (c_i + c_j) \\
& = w + c_j + w c_i + w c_j - c_i c_j - c_i^2 \\
\pd{c_i} u_i & = \pd{c_i} w c_i - \pd{c_i} c_i c_j - \pd{c_i} c_i^2 \\
& = w - c_j - 2 c_i \\
0 & \equiv w - c_j - 2 c_i \\
2 c_i & = w - c_j \\
c_i^\ast & = \frac{1}{2} \left( w - c_j \right)
\end{align*}

## Best Response Functions: Public good example

```{r public-good-b2, out.width = "70%", fig.width = 3, fig.height = 3, fig.align = "center"}
f = function(cj, w = 1) 0.5 * (w - cj)
A = seq(0, 1, 0.01)
d = data.frame(c1 = A, c2 = f(A))
ggplot(data = d, mapping = aes(x = c1, y = c2)) +
    geom_line(linewidth = 1, color = red) +
    geom_text(
        aes(x = 0.8, y = 0.25),
        label = "B[2](c[1])", parse = TRUE,
        color = red, fontface = "bold"
    ) +
    scale_x_continuous(breaks = c(0, 1/2, 1), labels = c(0, "w/2", "w"), limits = c(0, 1)) +
    scale_y_continuous(breaks = c(0, 1/2, 1), labels = c(0, "w/2", "w"), limits = c(0, 1)) +
    theme_classic()
```

## Best Response Functions: Public good example

```{r public-good-b1, out.width = "70%", fig.width = 3, fig.height = 3, fig.align = "center"}
d = data.frame(c2 = A, c1 = f(A))
ggplot(data = d, mapping = aes(x = c1, y = c2)) +
    geom_line(linewidth = 1, color = blue) +
    geom_text(
        aes(y = 0.8, x = 0.25),
        label = "B[1](c[2])", parse = TRUE,
        color = blue, fontface = "bold"
    ) +
    scale_x_continuous(breaks = c(0, 1/2, 1), labels = c(0, "w/2", "w"), limits = c(0, 1)) +
    scale_y_continuous(breaks = c(0, 1/2, 1), labels = c(0, "w/2", "w"), limits = c(0, 1)) +
    theme_classic()
```

## Best Response Functions: Public good example

```{r public-good-both, out.width = "70%", fig.width = 3, fig.height = 3, fig.align = "center"}
d = data.frame(c1 = c(A, f(A)), c2 = c(f(A), A), B = rep(c("B2", "B1"), each = length(A)))
pal = c(B1 = blue, B2 = red)
ggplot(data = d, mapping = aes(x = c1, y = c2, color = B)) +
    geom_line(linewidth = 1) +
    geom_point(aes(x = 1/3, y = 1/3), color = purp, size = 3) +
    geom_text(
        aes(x = 0.8, y = 0.25),
        label = "B[2](c[1])", parse = TRUE,
        color = red, fontface = "bold"
    ) +
    geom_text(
        aes(y = 0.8, x = 0.25),
        label = "B[1](c[2])", parse = TRUE,
        color = blue, fontface = "bold"
    ) +
    scale_x_continuous(breaks = c(0, 1/3, 1/2, 1), labels = c(0, "w/3", "w/2", "w"), limits = c(0, 1)) +
    scale_y_continuous(breaks = c(0, 1/3, 1/2, 1), labels = c(0, "w/3", "w/2", "w"), limits = c(0, 1)) +
    scale_color_manual(values = pal) +
    theme_classic() +
    theme(legend.position = "none")
```

## Best Response Functions: Public good example

\begin{align*}
c_i^\ast & = \frac{1}{2} \left( w - c_j^\ast \right) \\
\gray c_i^\ast & \gray = \frac{1}{2} \left( w - \frac{1}{2} \left( w - c_i^\ast \right) \right) \\
\gray c_i^\ast & \gray = w/4 + c_i^\ast/4 \\
\gray 4 c_i^\ast & \gray = w + c_i^\ast \\
\gray 3 c_i^\ast & \gray = w \\
\gray c_i^\ast & \gray = w / 3
\end{align*}

## Best Response Functions: Public good example

\begin{align*}
c_i^\ast & = \frac{1}{2} \left( w - c_j^\ast \right) \\
c_i^\ast & = \frac{1}{2} \left( w - \frac{1}{2} \left( w - c_i^\ast \right) \right) \\
\gray c_i^\ast & \gray = w/4 + c_i^\ast/4 \\
\gray 4 c_i^\ast & \gray = w + c_i^\ast \\
\gray 3 c_i^\ast & \gray = w \\
\gray c_i^\ast & \gray = w / 3
\end{align*}

## Best Response Functions: Public good example

\begin{align*}
c_i^\ast & = \frac{1}{2} \left( w - c_j^\ast \right) \\
c_i^\ast & = \frac{1}{2} \left( w - \frac{1}{2} \left( w - c_i^\ast \right) \right) \\
 c_i^\ast &  = w/4 + c_i^\ast/4 \\
 4 c_i^\ast &  = w + c_i^\ast \\
 3 c_i^\ast &  = w \\
 c_i^\ast &  = w / 3
\end{align*}
