---
title: "Intro to Formal Political Analysis:\\newline Strategic Games and Nash Equilibria \\newline in Pure Strategies"
author: "JBrandon Duck-Mayr"
date: "Spring 2023"
urlcolor: Blue
header-includes:
    - \newcommand{\setsep}{\setlength{\itemsep}{3pt}}
    - \newcommand{\setskip}{\setlength{\parskip}{3pt}}
    - \renewcommand{\tightlist}{\setsep\setskip}
    - \setlength{\abovedisplayskip}{-6pt}
    - \newcommand{\pd}[1]{\ensuremath{\frac{\partial}{\partial{#1}}}}
    - \newcommand{\gray}{\color{white}}
output:
    quack::presentation:
        toc: false
        incremental: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
quack::attach_common_packages()
```

## What is a "strategic game"?

. . .

::: {.block}
### Osborne, Definition 13.1
A strategic game consists of

\vspace*{3pt}

. . .

- a set of **players**
- for each player, a set of **actions**
- for each player, **preferences** over the set of action profiles
:::

. . .

&nbsp;

This is also called a "normal-form game"

## Example: The Prisoner's Dilemma

. . .

- Two suspects in a major crime are questioned separately
- The cops already have enough evidence to convict them of a minor crime but need a confession to convict on the major crime
- So if neither talks, they both get convicted of the minor crime but neither gets convicted of the major crime
- If only one of them talks, that prisoner will get a plea deal and serve no time while the other will be convicted of the major crime
- If both talk, they are both convicted of the major crime, but get a slightly reduced sentence for confessing

## The Prisoner's Dilemma as a normal form game

. . .

- Players: The two suspects
- Actions: Each player can stay quiet ($C$) or talk ($D$)
- Preferences:
  + From best to worst, Player 1 prefers:
    - $(D, C)$; 1 goes free and 2 is convicted of the major crime
    - $(C, C)$; both are only convicted of the minor crime only
    - $(D, D)$; both are convicted of the felony with a reduced sentence
    - $(C, D)$; 1 is convicted of the major crime and pays full freight
  + Player 2's ordering is $(C, D)$, $(C, C)$, $(D, D)$, and $(D, C)$.

. . .

&nbsp;

A utility function that represents Player 1's preferences is $$u_1(D, C) = 3,\ u_1(C, C) = 2,\ u_1(D, D) = 1,\ u_1(C, D) = 0$$

## Matrix representation of The Prisoner's Dilemma

. . .

```{r prisoners-dilemma-matrix}
library(kableExtra)
game = data.frame(
    C = c("$2, 2$", "$3, 0$"),
    D = c("$0, 3$", "$1, 1$"),
    row.names = c("C", "D")
)
kable(game, escape = FALSE, align = "c") %>% kable_styling(position = "center")
```

## Matrix representation of The Prisoner's Dilemma

I produced that for these slides using R Markdown with the code:

````{verbatim, echo = TRUE}
```{r prisoners-dilemma-matrix}
library(kableExtra)
game = data.frame(
    C = c("$2, 2$", "$3, 0$"),
    D = c("$0, 3$", "$1, 1$"),
    row.names = c("C", "D")
)
kable(game, escape = FALSE, align = "c") %>%
    kable_styling(position = "center")
```
````

## Matrix representation of The Prisoner's Dilemma

You could produce a similar payoff matrix in LaTeX with the code:

````{verbatim, lang = "latex", echo = TRUE}
\begin{game}{2}{2}
    & $C$       & $D$       \\
$C$ & $2, 2$    & $0, 3$    \\
$D$ & $3, 0$    & $1, 1$
\end{game}
````

(this requires you add `\usepackage{sgame}` to your preamble)

## Nash Equilibrium

. . .

::: {.block}
### Osborne, Definition 23.1
The action profile $a^\ast$ in a strategic game... is a **Nash equilibrium** if, for every player $i$ and every action $a_i$ of player $i$, $a^\ast$ is at least as good according to player $i$'s preferences as the action profile $(a_i, a_{-i}^\ast)$ in which player $i$ chooses $a_i$ while every other player $j$ chooses $a_j^\ast$. Equivalently, for every player $i$,

$$u_i\left(a^\ast\right) \geq u_i\left(a_i, a_{-i}^\ast\right) \text{ for every action } a_i \text{ of player } i,$$

where $u_i$ is a payoff function that represents player $i$'s preferences.
:::

. . .

In other words, a Nash equilibrium is an action profile where no player can gain by unilateral deviation

## Nash equilibrium in The Prisoner's Dilemma

\begin{table}
\begin{tabular}{lcc}
& C & D \\
C & 2, 2 & 0, \textbf<4->{3} \\
D & \textbf<2->{3}, 0 & \textbf<3->{1}, \textbf<5->{1}
\end{tabular}
\end{table}

\phantom{
\parbox{\linewidth}{
\vspace*{\baselineskip}
\begin{itemize}
\tightlist
\item $(D, D)$ is the only profile where neither can gain by deviating
\item It is therefore the unique Nash equilibrium
\item Notice they'd both be better off in the action profile $(C, C)$
\item But Nash equilibrium is about unilateral deviation
\end{itemize}
}
}

## Nash equilibrium in The Prisoner's Dilemma

\begin{table}
\begin{tabular}{lcc}
& C & D \\
C & \textcolor<3->{BurntOrange}{\textbf<3->{2, 2}} & 0, \textbf{3} \\
D & \textbf{3}, 0 & \textbf{1}, \textbf{1}
\end{tabular}
\end{table}

&nbsp;

- $(D, D)$ is the only profile where neither can gain by deviating
- It is therefore the unique Nash equilibrium
- Notice they'd both be better off in the action profile $(C, C)$
- But Nash equilibrium is about *unilateral* deviation

## Domination

\only<2>{
\begin{block}{Osborne, Definition 45.1}
In a strategic game... player $i$'s action $a_i''$ \textbf{strictly dominates} her action $a_i'$ if {\small $$ u_i\left(a_i'', a_{-i}\right) > u_i\left(a_i', a_{-i}\right) \text{ for every list } a_{-i} \text{ of the other players' actions}, $$}

where $u_i$ is a payoff function that represents player $i$'s preferences.
We say that the action $a_i'$ is \textbf{strictly dominated}.
\end{block}
}

\only<3>{
\begin{block}{Osborne, Definition 46.1}
In a strategic game... player $i$'s action $a_i''$ \textbf{weakly dominates} her action $a_i'$ if {\small $$ u_i\left(a_i'', a_{-i}\right) \geq u_i\left(a_i', a_{-i}\right) \text{ for every list } a_{-i} \text{ of the other players' actions}, $$}

and {\small $$ u_i\left(a_i'', a_{-i}\right) > u_i\left(a_i', a_{-i}\right) \text{ for some list } a_{-i} \text{ of the other players' actions}, $$}

where $u_i$ is a payoff function that represents player $i$'s preferences.
We say that the action $a_i'$ is \textbf{weakly dominated}.
\end{block}
}

\only<4>{
In other words,

\begin{enumerate}
\tightlist
\item Strict domination: Action A strictly dominates action B if the player strictly prefers all action profiles where they play A to all action profiles where they play B
\item Weak domination: A weakly dominates B if we change "strictly prefers" to "weakly prefers" \textbf{and} require the preference is strict in at least one action profile
\end{enumerate}
}

## Domination in The Prisoner's Dilemma

\begin{table}
\begin{tabular}{lcc}
& C & D \\
C & 2, 2 & 0, \textbf<4->{3} \\
D & \textbf<2->{3}, 0 & \textbf<3->{1}, \textbf<5->{1}
\end{tabular}
\end{table}

\onslide<6>{
\vspace*{\baselineskip}
Whether Player 1 plays $C$ or $D$, Player 2 strictly prefers to play $D$.
Therefore, for Player 2, $D$ \textbf{strictly dominates} $C$.
}

## Another example of domination

\begin{table}
\begin{tabular}{lcc}
  & L & R \\
T & \textcolor<3>{Blue}{\textbf<2-3>{1}} & \textcolor<3>{Blue}{\textbf<2-3>{0}} \\
M & \textcolor<3>{BurntOrange}{\textcolor<5>{Blue}{\textbf<2-5>{2}}} & \textcolor<3>{BurntOrange}{\textcolor<5>{Blue}{\textbf<2-5>{1}}} \\
B & \textcolor<5>{BurntOrange}{\textbf<4-5>{3}} & \textcolor<5>{BurntOrange}{\textbf<4-5>{2}}
\end{tabular}
\end{table}

\onslide<6>{
\vspace*{\baselineskip}
For Player 1, $M$ \textbf{strictly dominates} $T$, and $B$ \textbf{strictly dominates} $M$.
}

## Example: Bach or Stravinski?

Two people want to go out on Friday.
There's a concert playing music by Bach and a concert playing music by Stravinski.
One person prefers Bach while the other prefers Stravinski.
For both of them, the worst option is going to a concert alone.

## Example: Bach or Stravinski?

- Players: The two people
- Actions: Each player can go to the Bach concert ($B$) or the Stravinski concert ($S$)
- Preferences:
  + From best to worst, Player 1 prefers $(B, B), (S, S), (B, S), (S, B)$
  + From best to worst, Player 2 prefers $(S, S), (B, B), (B, S), (S, B)$
  + Assuming they don't care about which concert they go to if alone, a utility function that represents Player 1's preferences is $$u_1(B, B) = 2,\ u_1(S, S) = 1,\ u_1(B, S) = u_1(S, B) = 0$$

## Example: Bach or Stravinski?

\begin{table}
\begin{tabular}{lcc}
& B & S \\
B & \textbf<2->{2}, \textbf<4->{1} & 0, 0 \\
S & 0, 0 & \textbf<3->{1}, \textbf<5->{2}
\end{tabular}
\end{table}

\onslide<6->{
In both the $(B, B)$ and $(S, S)$ profiles, neither player can gain from unilateral deviation.

Therefore, both action profiles are a Nash equilibrium.
}

\onslide<7>{
Does any of Player 1's actions weakly or strictly dominate any of their other actions?
How about for Player 2?
}

## Example: Stag Hunt

::: {.block}
### Osborne, page 20 in the 9th printing
A sentence in *Discourse on the origin and foundations of inequality among men* (1755) by the philosopher Jean-Jacques Rousseau discusses a group of hunters who wish to catch a stag... They will succeed if they all remain sufficiently attentive, but each is tempted to desert her post and catch a hare.
:::

## Example: Stag Hunt

- Players: The hunters
- Actions: Each can remain attentive to the stag ($S$) or desert and catch a hare ($H$)
- Preferences: Each best prefers a profile where every player chooses $S$, next prefers any profile where they personally choose $H$, and lastly prefer any profile where they choose $S$ and at least one other player chooses $H$

## Example: Stag Hunt

A matrix representation of the two-person stag hunt is

\begin{table}
\begin{tabular}{lcc}
& S & H \\
S & \textbf<2->{2}, \textbf<4->{2} & 0, 1 \\
H & 1, 0 & \textbf<3->{1}, \textbf<5->{1}
\end{tabular}
\end{table}

\onslide<6->{
In both the $(S, S)$ and $(H, H)$ profiles, neither player can gain from unilateral deviation.

Therefore, both action profiles are a Nash equilibrium.
}

\onslide<7>{
Does any of Player 1's actions weakly or strictly dominate any of their other actions?
How about for Player 2?
}

## Example: Matching Pennies

Two people simultaneously show each other a side of a penny.
If they show each other the same side, Player 2 pays Player 1 $1, while if they show each other different sides, Player 1 pays Player 2 $1.

## Example: Matching Pennies

- Players: The two people
- Actions: Each player can show heads ($H$) or show tails ($T$)
- Preferences:
  + From best to worst, Player 1 prefers $(H, H)$ or $(T, T)$, then $(H, T)$ or $(T, H)$
  + From best to worst, Player 2 prefers $(H, T)$ or $(T, H)$, then $(H, H)$ or $(T, T)$
  + We can represent their preferences by how much money they gain or lose; $$ u_1(H, H) = u_1(T, T) = 1,\ u_1(H, T) = u_1(T, H) = -1 $$

## Example: Matching Pennies

\begin{table}
\begin{tabular}{lcc}
& H & T \\
H & \textbf<2->{1}, -1 & -1, \textbf<4->{1} \\
T & -1, \textbf<5->{1} & \textbf<3->{1}, -1
\end{tabular}
\end{table}

\onslide<6->{
There is no pure strategy Nash equilibrium in the Matching Pennies game.
}

\onslide<7>{
Does any of Player 1's actions weakly or strictly dominate any of their other actions?
How about for Player 2?
}

## Example: Voting

Consider the following scenario:

- Two candidates, $A$ and $B$, run for office
- There are $N$ voters, with $N$ odd; some of them (a majority) prefer candidate $A$ to candidate $B$, and the rest prefer $B$ to $A$.
- There is an election with mandatory voting; the candidate with the most votes wins

## Example: Voting

- Players: The $N$ voters
- Actions: Each can vote for $A$ or vote for $B$
- Preferences:
  + All players are indifferent between action profiles where $A$ wins
  + All players are indifferent between action profiles where $B$ wins
  + Players who prefer candidate $A$ prefer any action profile where $A$ wins to any action profile where $B$ wins, and vice versa for voters who prefer candidate $B$

## Example: Voting

Questions:

- Do the players have an action that dominates their other action?
- Is there any Nash equilibrium where candidate $A$ wins the election?
- Is there any Nash equilibrium where candidate $B$ wins the election?

## Dominance Solvable Games

. . .

- A game is **dominance solvable** if iterated elimination of dominated actions leaves only one actions left for each player
- If you are eliminating only strictly dominated strategies, the remaining action profile is the unique Nash equilibrium to the game (IESDS / IDSDS)
  + Order of elimination does **not** matter
- If you are also eliminating weakly dominated strategies, the remaining action profile will be a Nash equilibrium, but you may have eliminated other Nash equilibria (IEWDS / IDWDS)
  + Order of elimination **does** matter

## Dominance Solvable Games: Examples

\begin{figure}
\centering
The Prisoner's Dilemma
\end{figure}
\begin{table}
\begin{tabular}{lcc}
& \textcolor<5>{LightGray}{C} & \textbf<5>{D} \\
\textcolor<3->{LightGray}{C} &
\textcolor<2>{Blue}{\textcolor<3->{LightGray}{\textbf<2>{2}}}\textcolor<3->{LightGray}{,} \textcolor<3->{LightGray}{2} &
\textcolor<2>{Blue}{\textcolor<3->{LightGray}{\textbf<2>{0}}}\textcolor<3->{LightGray}{,} \textcolor<3->{LightGray}{3} \\
\textbf<5>{D} &
\textcolor<2>{DarkGreen}{\textcolor<5>{LightGray}{\textbf<2>{3}}}\textcolor<5>{LightGray}{,} \textcolor<4>{Blue}{\textcolor<5>{LightGray}{\textbf<4>{0}}} &
\textcolor<2>{DarkGreen}{\textbf<2, 5>{1}}\textbf<5>{, }\textcolor<4>{DarkGreen}{\textbf<4, 5>{1}}
\end{tabular}
\end{table}

## Dominance Solvable Games: Examples

\begin{figure}
\centering
Bach or Stravinski
\end{figure}
\begin{table}
\begin{tabular}{lcc}
& B & S \\
B & 2, 1 & 0, 0 \\
S & 0, 0 & 1, 2
\end{tabular}
\end{table}

## Dominance Solvable Games: Examples

\begin{table}
\begin{tabular}{lccc}
& L & C & R \\
T & 0, 0 & 1, 0 & 1, 1 \\
M & 1, 0 & 1, 1 & 3, 2 \\
B & 1, 0 & 2, 1 & 2, 2
\end{tabular}
\end{table}

## Dominance Solvable Games: Examples

\begin{table}
\begin{tabular}{lccc}
& L & C & R \\
T & 0, 0 & 1, 0 & 1, 1 \\
M & 1, 1 & 1, 1 & 3, 2 \\
B & 1, 1 & 2, 1 & 2, 2
\end{tabular}
\end{table}

## More Voting: The Hotelling-Downs Model

. . .

Consider the following model of elections:

- Two candidates seek election & *only* care about winning office
- (They prefer winning to tying and prefer tying to losing)
- Suppose there is a unidimensional policy space
- The candidates run for office on a platform (a point)
- There are a continuum of voters each with an ideal point
- They vote for the candidate closest to their ideal point
- (If multiple platforms are equidistant from a point, the candidates evenly split the relevant voters)

## More Voting: The Hotelling-Downs Model

Questions:

- Who are the players?
- Are you already familiar with the game's solution?
- How would you find the solution to this game?

## Even More Voting! The Calvert-Wittman Model

- What if the candidates cared about policy rather than winning?
- (These are sometimes called "citizen-candidates")
- Does this change the Nash equilibrium? (Focus on the case where the two candidates' ideal points are on opposite sides of the median)

## Even More Voting!!!

We could think about lots more extensions or tweaks:

. . .

- What about asymmetric voter preferences?
- What about "valence"?
- What about stochastic voting?
- What about multiple dimensions?
- Etc, etc, etc

## Best Response Functions

The \textbf{best response function} of player $i$ is $$
B_i\left(a_{-i}\right) = \{ a_i \in A_i : u_i\left(a_i, a_{-i}\right) \geq u_i\left(a_i', a_{-i}\right) \forall\ a_i' \in A_i \}$$

\invisible{
\begin{block}{Osborne Proposition 36.1}
The action profile $a^\ast$ is a Nash equilibrium of a strategic game... if and only if every player's action is a best response to the other player's actions:

$$
a_i^\ast \text{ is in } B_i\left(a_{-i}^\ast\right) \text{ for every player } i.
$$
\end{block}
}

## Best Response Functions

The \textbf{best response function} of player $i$ is $$
\mathbf{B_i\left(a_{-i}\right)} = \{ a_i \in A_i : u_i\left(a_i, a_{-i}\right) \geq u_i\left(a_i', a_{-i}\right) \forall\ a_i' \in A_i \}$$

\invisible{
\begin{block}{Osborne Proposition 36.1}
The action profile $a^\ast$ is a Nash equilibrium of a strategic game... if and only if every player's action is a best response to the other player's actions:

$$
a_i^\ast \text{ is in } B_i\left(a_{-i}^\ast\right) \text{ for every player } i.
$$
\end{block}
}

## Best Response Functions

The \textbf{best response function} of player $i$ is $$
B_i\left(a_{-i}\right) = \textcolor{BurntOrange}{\Bigl\{} a_i \in A_i : u_i\left(a_i, a_{-i}\right) \geq u_i\left(a_i', a_{-i}\right) \forall\ a_i' \in A_i \textcolor{BurntOrange}{\Bigr\}}$$

\invisible{
\begin{block}{Osborne Proposition 36.1}
The action profile $a^\ast$ is a Nash equilibrium of a strategic game... if and only if every player's action is a best response to the other player's actions:

$$
a_i^\ast \text{ is in } B_i\left(a_{-i}^\ast\right) \text{ for every player } i.
$$
\end{block}
}

## Best Response Functions

The \textbf{best response function} of player $i$ is $$
B_i\left(a_{-i}\right) = \{ a_i \in A_i : u_i\left(a_i, a_{-i}\right) \geq u_i\left(a_i', a_{-i}\right) \forall\ a_i' \in A_i \}$$

::: {.block}
### Osborne Proposition 36.1
The action profile $a^\ast$ is a Nash equilibrium of a strategic game... if and only if every player's action is a best response to the other player's actions:

$$
a_i^\ast \text{ is in } B_i\left(a_{-i}^\ast\right) \text{ for every player } i.
$$
:::

## Best Response Functions: Voting example

- Let's revisit the Hotelling-Downs model of elections
- Candidate 2's platform $x_2$ could be
  + $x_2 < m$
  + $x_2 = m$
  + $x_2 > m$

\onslide<6>{
$$
B_1\left(x_2\right) =
\begin{cases}
\left\{ x_1 : x_2 < x_1 < 2m - x_2 \right\} & \text{ if } x_2 < m \\
\left\{ m \right\} & \text{ if } x_2 = m \\
\left\{ x_1 : 2m - x_2 < x_1 < x_2 \right\} & \text{ if } x_2 < m
\end{cases}
$$
}

## Best Response Functions: Voting example

```{r voting-br1, out.width = "70%", fig.width = 3, fig.height = 3, fig.align = "center"}
dots = data.frame(x1 = c(0, 0, 1, 1), x2 = c(0, 1, 0, 1))
segs = data.frame(x1 = c(0, 0), x2 = c(0, 1), x11 = c(1, 1), x21 = c(0, 1))
shp1 = data.frame(x1 = c(0, 0.5, 1), x2 = c(1, 0.5, 1))
shp2 = data.frame(x1 = c(0, 0.5, 1), x2 = c(0, 0.5, 0))
blue = "dodgerblue4"
ggplot(data = dots, mapping = aes(x = x1, y = x2)) +
    geom_polygon(data = shp1, fill = blue, alpha = 1/2) +
    geom_polygon(data = shp2, fill = blue, alpha = 1/2) +
    geom_text(
        aes(x = 0.5, y = 0.75),
        label = "B[1](x[2])", parse = TRUE,
        color = blue, fontface = "bold"
    ) +
    geom_point(aes(x = 0.5, y = 0.5), shape = 19, color = blue, size = 2) +
    geom_segment(
        mapping = aes(xend = x11, yend = x21), data = segs,
        color = blue, linewidth = 1
    ) +
    geom_point(shape = 21, color = blue, fill = "white", size = 2) +
    scale_x_continuous(breaks = 0.5, labels = "m") +
    scale_y_continuous(breaks = 0.5, labels = "m") +
    theme_classic()
```

## Best Response Functions: Voting example

```{r voting-br2, out.width = "70%", fig.width = 3, fig.height = 3, fig.align = "center"}
seg2 = data.frame(x2 = c(0, 0), x1 = c(0, 1), x21 = c(1, 1), x11 = c(0, 1))
shp3 = data.frame(x2 = c(0, 0.5, 1), x1 = c(1, 0.5, 1))
shp4 = data.frame(x2 = c(0, 0.5, 1), x1 = c(0, 0.5, 0))
red  = "firebrick3"
ggplot(data = dots, mapping = aes(x = x1, y = x2)) +
    geom_polygon(data = shp3, fill = red,  alpha = 1/2) +
    geom_polygon(data = shp4, fill = red,  alpha = 1/2) +
    geom_text(
        aes(y = 0.5, x = 0.75),
        label = "B[2](x[1])", parse = TRUE,
        color = red, fontface = "bold"
    ) +
    geom_point(aes(x = 0.5, y = 0.5), shape = 19, color = red, size = 2) +
    geom_segment(
        mapping = aes(xend = x11, yend = x21), data = seg2,
        color = red, linewidth = 1
    ) +
    geom_point(shape = 21, color = red, fill = "white", size = 2) +
    scale_x_continuous(breaks = 0.5, labels = "m") +
    scale_y_continuous(breaks = 0.5, labels = "m") +
    theme_classic()
```

## Best Response Functions: Voting example

```{r voting-both-br, out.width = "70%", fig.width = 3, fig.height = 3, fig.align = "center"}
purp = "#6e3a58"
ggplot(data = dots, mapping = aes(x = x1, y = x2)) +
    geom_polygon(data = shp1, fill = blue, alpha = 1/2) +
    geom_polygon(data = shp2, fill = blue, alpha = 1/2) +
    geom_polygon(data = shp3, fill = red,  alpha = 1/2) +
    geom_polygon(data = shp4, fill = red,  alpha = 1/2) +
    geom_text(
        aes(x = 0.5, y = 0.75),
        label = "B[1](x[2])", parse = TRUE,
        color = blue, fontface = "bold"
    ) +
    geom_text(
        aes(y = 0.5, x = 0.75),
        label = "B[2](x[1])", parse = TRUE,
        color = red, fontface = "bold"
    ) +
    geom_point(aes(x = 0.5, y = 0.5), shape = 19, color = purp, size = 2) +
    geom_segment(
        mapping = aes(xend = x11, yend = x21), data = segs,
        color = blue, linewidth = 1
    ) +
    geom_segment(
        mapping = aes(xend = x11, yend = x21), data = seg2,
        color = red, linewidth = 1
    ) +
    geom_point(shape = 21, color = purp, fill = "white", size = 2) +
    scale_x_continuous(breaks = 0.5, labels = "m") +
    scale_y_continuous(breaks = 0.5, labels = "m") +
    theme_classic()
```

## Best Response Functions: Public good example

- Now let's consider a public good problem (Osborne Ex. 44.1)
- Two people with wealth $w$ each choose a contribution level $c_i$ to the public good
- Suppose they value the public good itself ($c_1 + c_2$) as well as their own private consumption ($w - c_i$) and value their private consumption more as the amount of the public good increases

\onslide<4>{
\begin{align*}
u_i\left(c_1, c_2\right)
& = c_1 + c_2 + w - c_i + (w - c_i) (c_1 + c_2) \\
& = w + c_j + (w - c_i) (c_1 + c_2)
\end{align*}
}

## Best Response Functions: Public good example

::: {.nonincremental}

- Now let's consider a public good problem (Osborne Ex. 44.1)
- Two people with wealth $w$ each choose a contribution level $c_i$ to the public good
- Suppose they value the public good itself ($c_1 + c_2$) as well as their own private consumption ($w - c_i$) and value their private consumption more as the amount of the public good increases: $$ u_i\left(c_1, c_2\right) = w + c_j + (w - c_i) (c_1 + c_2) $$
- How would you find the players' best response functions?

:::

## Best Response Functions: Public good example

\begin{align*}
u_i\left(c_i, c_j\right) & = w + c_j + (w - c_i) (c_i + c_j) \\
& = w + c_j + w c_i + w c_j - c_i c_j - c_i^2 \\
\gray\pd{c_i} u_i
&\gray = \pd{c_i} w c_i - \pd{c_i} c_i c_j - \pd{c_i} c_i^2 \\
&\gray = w - c_j - 2 c_i \\
\gray 0 &\gray \equiv w - c_j - 2 c_i \\
\gray 2 c_i &\gray = w - c_j \\
\gray c_i^\ast &\gray = \frac{1}{2} \left( w - c_j \right)
\end{align*}

## Best Response Functions: Public good example

\begin{align*}
u_i\left(c_i, c_j\right) & = w + c_j + (w - c_i) (c_i + c_j) \\
& = w + c_j + w c_i + w c_j - c_i c_j - c_i^2 \\
\pd{c_i} u_i & = \pd{c_i} w c_i - \pd{c_i} c_i c_j - \pd{c_i} c_i^2 \\
& = w - c_j - 2 c_i \\
\gray 0 & \gray\equiv w - c_j - 2 c_i \\
\gray 2 c_i & \gray = w - c_j \\
\gray c_i^\ast & \gray = \frac{1}{2} \left( w - c_j \right)
\end{align*}

## Best Response Functions: Public good example

\begin{align*}
u_i\left(c_i, c_j\right) & = w + c_j + (w - c_i) (c_i + c_j) \\
& = w + c_j + w c_i + w c_j - c_i c_j - c_i^2 \\
\pd{c_i} u_i & = \pd{c_i} w c_i - \pd{c_i} c_i c_j - \pd{c_i} c_i^2 \\
& = w - c_j - 2 c_i \\
0 & \equiv w - c_j - 2 c_i \\
2 c_i & = w - c_j \\
c_i^\ast & = \frac{1}{2} \left( w - c_j \right)
\end{align*}

## Best Response Functions: Public good example

```{r public-good-b2, out.width = "70%", fig.width = 3, fig.height = 3, fig.align = "center"}
f = function(cj, w = 1) 0.5 * (w - cj)
A = seq(0, 1, 0.01)
d = data.frame(c1 = A, c2 = f(A))
ggplot(data = d, mapping = aes(x = c1, y = c2)) +
    geom_line(linewidth = 1, color = red) +
    geom_text(
        aes(x = 0.8, y = 0.25),
        label = "B[2](c[1])", parse = TRUE,
        color = red, fontface = "bold"
    ) +
    scale_x_continuous(breaks = c(0, 1/2, 1), labels = c(0, "w/2", "w"), limits = c(0, 1)) +
    scale_y_continuous(breaks = c(0, 1/2, 1), labels = c(0, "w/2", "w"), limits = c(0, 1)) +
    theme_classic()
```

## Best Response Functions: Public good example

```{r public-good-b1, out.width = "70%", fig.width = 3, fig.height = 3, fig.align = "center"}
d = data.frame(c2 = A, c1 = f(A))
ggplot(data = d, mapping = aes(x = c1, y = c2)) +
    geom_line(linewidth = 1, color = blue) +
    geom_text(
        aes(y = 0.8, x = 0.25),
        label = "B[1](c[2])", parse = TRUE,
        color = blue, fontface = "bold"
    ) +
    scale_x_continuous(breaks = c(0, 1/2, 1), labels = c(0, "w/2", "w"), limits = c(0, 1)) +
    scale_y_continuous(breaks = c(0, 1/2, 1), labels = c(0, "w/2", "w"), limits = c(0, 1)) +
    theme_classic()
```

## Best Response Functions: Public good example

```{r public-good-both, out.width = "70%", fig.width = 3, fig.height = 3, fig.align = "center"}
d = data.frame(c1 = c(A, f(A)), c2 = c(f(A), A), B = rep(c("B2", "B1"), each = length(A)))
pal = c(B1 = blue, B2 = red)
ggplot(data = d, mapping = aes(x = c1, y = c2, color = B)) +
    geom_line(linewidth = 1) +
    geom_point(aes(x = 1/3, y = 1/3), color = purp, size = 3) +
    geom_text(
        aes(x = 0.8, y = 0.25),
        label = "B[2](c[1])", parse = TRUE,
        color = red, fontface = "bold"
    ) +
    geom_text(
        aes(y = 0.8, x = 0.25),
        label = "B[1](c[2])", parse = TRUE,
        color = blue, fontface = "bold"
    ) +
    scale_x_continuous(breaks = c(0, 1/3, 1/2, 1), labels = c(0, "w/3", "w/2", "w"), limits = c(0, 1)) +
    scale_y_continuous(breaks = c(0, 1/3, 1/2, 1), labels = c(0, "w/3", "w/2", "w"), limits = c(0, 1)) +
    scale_color_manual(values = pal) +
    theme_classic() +
    theme(legend.position = "none")
```

## Best Response Functions: Public good example

\begin{align*}
c_i^\ast & = \frac{1}{2} \left( w - c_j^\ast \right) \\
\gray c_i^\ast & \gray = \frac{1}{2} \left( w - \frac{1}{2} \left( w - c_i^\ast \right) \right) \\
\gray c_i^\ast & \gray = w/4 + c_i^\ast/4 \\
\gray 4 c_i^\ast & \gray = w + c_i^\ast \\
\gray 3 c_i^\ast & \gray = w \\
\gray c_i^\ast & \gray = w / 3
\end{align*}

## Best Response Functions: Public good example

\begin{align*}
c_i^\ast & = \frac{1}{2} \left( w - c_j^\ast \right) \\
c_i^\ast & = \frac{1}{2} \left( w - \frac{1}{2} \left( w - c_i^\ast \right) \right) \\
\gray c_i^\ast & \gray = w/4 + c_i^\ast/4 \\
\gray 4 c_i^\ast & \gray = w + c_i^\ast \\
\gray 3 c_i^\ast & \gray = w \\
\gray c_i^\ast & \gray = w / 3
\end{align*}

## Best Response Functions: Public good example

\begin{align*}
c_i^\ast & = \frac{1}{2} \left( w - c_j^\ast \right) \\
c_i^\ast & = \frac{1}{2} \left( w - \frac{1}{2} \left( w - c_i^\ast \right) \right) \\
 c_i^\ast &  = w/4 + c_i^\ast/4 \\
 4 c_i^\ast &  = w + c_i^\ast \\
 3 c_i^\ast &  = w \\
 c_i^\ast &  = w / 3
\end{align*}
