---
title: "Intro to Formal Political Analysis:\\newline Mixed Strategies in Strategic Games"
author: "JBrandon Duck-Mayr"
date: "Spring 2023"
urlcolor: Blue
header-includes:
    - \newcommand{\setsep}{\setlength{\itemsep}{3pt}}
    - \newcommand{\setskip}{\setlength{\parskip}{3pt}}
    - \renewcommand{\tightlist}{\setsep\setskip}
    - \setlength{\abovedisplayskip}{-6pt}
    - \newcommand{\pd}[1]{\ensuremath{\frac{\partial}{\partial{#1}}}}
    - \newcommand{\gray}{\color{white}}
output:
    quack::presentation:
        toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
quack::attach_common_packages()
```

## Recap

::: nonincremental
- A strategic or normal-form game consists of
  + Players
  + For each player, a set of actions
  + For each player, preferences over the set of action profiles
- Nash equilibrium: quality of an action profile
  + Action profile is NE if no player can gain from unilateral deviation
- Domination: comparison between two actions of one player
  + Strict domination: Action A strictly dominates action B if the player strictly prefers all action profiles where they play A to all action profiles where they play B
  + Weak domination: A weakly dominates B if we change "strictly prefers" to "weakly prefers" **and** require the preference is strict in at least one action profile
:::

## Recap

::: nonincremental
- A game is dominance solvable if iterated elimination of dominated actions leaves only one actions left for each player
- A player's best response function tells you which of their actions are at least as good as any of their other available actions according to their preferences, given a combination of actions of every other player
- An action profile is a Nash equilibrium of a strategic game if and only if every player's action is a best response to the other player's actions
:::

## Mixed Strategies

Now we will introduce "mixed strategies", or allowing the players to randomize

::: {.block}
### Osborne Definition 107.1
A **mixed strategy** of a player in a strategic game is a probability distribution over the player's actions.
:::

## Mixed Strategy Nash Equilibrium

::: {.block}
### Osborne Proposition 116.2
A mixed strategy profile $\alpha^\ast$ in a strategic game with vNM preferences in which each player has finitely many actions is a mixed strategy Nash equilibrium if and only if, for each player $i$,

&nbsp;

- the expected payoff, given $\alpha_{-i}^\ast$, to every action to which $\alpha_i^\ast$ assigns positive probability is the same
- The expected payoff, given $\alpha_{-i}^\ast$, to every action to which $\alpha_i^\ast$ assigns zero probability is at most the expected payoff to any action to which $\alpha_i^\ast$ assigns positive probability.

&nbsp;

Each player's expected payoff in an equilibrium is her expected payoff to any of her actions that she uses with positive probability
:::

## Example: Matching Pennies

\begin{table}
\begin{tabular}{lcc}
& H & T \\
H & 1, -1 & -1, 1 \\
T & -1, 1 & 1, -1
\end{tabular}
\end{table}

. . .

Let $p$ be the probability player 1 chooses $H$ and $q$ be the probability 2 chooses $H$.

. . .

Then

\begin{align*}
EU_1(H) & =  q - (1 - q) \\
EU_1(T) & = -q + (1 - q).
\end{align*}

## Example: Matching Pennies

Player 1 can mix over $H$ and $T$ if

\begin{align*}
EU_1(H) & = EU_1(T) \\
q - (1 - q) & = -q + (1 - q) \\
q - 1 + q & = -q + 1 - q \\
2q - 1 & = 1 - 2q \\
4q & = 2 \\
q & = 1/2.
\end{align*}

## Example: Matching Pennies

Similarly,

\begin{align*}
EU_2(H) & = -p + (1 - p) \\
EU_2(T) & =  p - (1 - p),
\end{align*}

and player 2 can mix over $H$ and $T$ if

\begin{align*}
EU_2(H) & = EU_2(T) \\
-p + (1 - p) & = p - (1 - p) \\
1 - 2p & = 2p - 1 \\
2 & = 4p \\
1/2 & = p.
\end{align*}

## Example: Matching Pennies

The players' best response functions, then, are

. . .

$$
BR_1(q) =
\begin{cases}
       p = 1 & \text{ if } q > 1/2 \\
p \in [0, 1] & \text{ if } q = 1/2 \\
       p = 0 & \text{ if } q < 1/2
\end{cases},
$$

and

$$
BR_2(p) =
\begin{cases}
       q = 1 & \text{ if } p < 1/2 \\
q \in [0, 1] & \text{ if } p = 1/2 \\
       q = 0 & \text{ if } p > 1/2
\end{cases}.
$$

## Example: Matching Pennies


```{r mp-br, out.width = "70%", fig.width = 3, fig.height = 3, fig.align = "center"}
d = data.frame(
    p = c(0, 1/2, 1/2, 1, 0,   0,   1, 1),
    q = c(1,   1,   0, 0, 0, 1/2, 1/2, 1),
    Label = c(rep("BR2", 4), rep("BR1", 4))
)
e = data.frame(
    p = c(1/2),
    q = c(1/2),
    Label = "NE"
)
pal = c(BR1 = "dodgerblue4", BR2 = "firebrick3", NE = "maroon4")
ggplot(data = d, mapping = aes(x = p, y = q, color = Label)) +
    geom_line(linewidth = 1) +
    geom_point(data = e, size = 3) +
    geom_text(
        mapping = aes(x = 0.2, y = 0.4, label = "BR[1](q)"),
        color = "dodgerblue4", parse = TRUE
    ) +
    geom_text(
        mapping = aes(x = 0.65, y = 0.8, label = "BR[2](p)"),
        color = "firebrick3", parse = TRUE
    ) +
    scale_color_manual(values = pal) +
    theme_bw() +
    theme(legend.position = "none")
```

## Example: Bach or Stravinski

\begin{table}
\begin{tabular}{lcc}
& B & S \\
B & 2, 1 & 0, 0 \\
S & 0, 0 & 1, 2
\end{tabular}
\end{table}

. . .

Let $p$ be the probability Player 1 plays $B$ and $q$ be the probability Player 2 plays $B$.

. . .

Then

\begin{align*}
EU_1(B) & = 2q \\
EU_1(S) & = 1 - q.
\end{align*}

## Example: Bach or Stravinski

Player 1 can mix over $B$ and $S$ if

\begin{align*}
EU_1(B) & = EU_1(S) \\
2q & = 1 - q \\
3q & = 1 \\
q & = 1/3.
\end{align*}

## Example: Bach or Stravinski

Similarly,

\begin{align*}
EU_2(B) & = p \\
EU_2(S) & = 2(1 - p),
\end{align*}

and player 2 can mix over $B$ and $S$ if

\begin{align*}
EU_2(B) & = EU_2(S) \\
p & = 2(1 - p) \\
p & = 2 - 2p \\
3p & = 2 \\
p & = 2/3.
\end{align*}

## Example: Bach or Stravinski

The players' best response functions, then, are

. . .

$$
BR_1(q) =
\begin{cases}
       p = 1 & \text{ if } q > 1/3 \\
p \in [0, 1] & \text{ if } q = 1/3 \\
       p = 0 & \text{ if } q < 1/3
\end{cases},
$$

and

$$
BR_2(p) =
\begin{cases}
       q = 1 & \text{ if } p > 2/3 \\
q \in [0, 1] & \text{ if } p = 2/3 \\
       q = 0 & \text{ if } p < 2/3
\end{cases}.
$$

## Example: Bach or Stravinski

```{r b-s-br, out.width = "70%", fig.width = 3, fig.height = 3, fig.align = "center"}
d = data.frame(
    p = c(0, 2/3, 2/3, 1, 0,   0,   1, 1),
    q = c(0,   0,   1, 1, 0, 1/3, 1/3, 1),
    Label = c(rep("BR2", 4), rep("BR1", 4))
)
e = data.frame(
    p = c(0, 2/3, 1),
    q = c(0, 1/3, 1),
    Label = "NE"
)
pal = c(BR1 = "dodgerblue4", BR2 = "firebrick3", NE = "maroon4")
ggplot(data = d, mapping = aes(x = p, y = q, color = Label)) +
    geom_line(linewidth = 1) +
    geom_point(data = e, size = 3) +
    geom_text(
        mapping = aes(x = 0.2, y = 0.4, label = "BR[1](q)"),
        color = "dodgerblue4", parse = TRUE
    ) +
    geom_text(
        mapping = aes(x = 0.5, y = 0.8, label = "BR[2](p)"),
        color = "firebrick3", parse = TRUE
    ) +
    scale_color_manual(values = pal) +
    theme_bw() +
    theme(legend.position = "none")
```

## Example: Bach or Stravinski

So

$$
(p, q) \in \{ (0, 0), (2/3, 1/3), (1, 1) \}
$$

are all Nash equilibria to the Bach or Stravinski game

## Example: Stag Hunt

\begin{table}
\begin{tabular}{lcc}
& S & H \\
S & 2, 2 & 0, 1 \\
H & 1, 0 & 1, 1
\end{tabular}
\end{table}

## Domination and Mixed Strategies

## Example 

\begin{table}
\begin{tabular}{lccc}
& L & C & R \\
T & 5, 1 & 1, 4 & 1, 0 \\
M & 3, 2 & 0, 0 & 3, 5 \\
B & 4, 3 & 4, 4 & 0, 3
\end{tabular}
\end{table}

\onslide<2>{
$$
EU_2(0, 1/2, 1/2) =
\begin{cases}
2   & \text{ if } 1 \text{ chooses } T \\
2.5 & \text{ if } 1 \text{ chooses } M \\
3.5 & \text{ if } 1 \text{ chooses } B
\end{cases}
$$

}

## Example 

\begin{table}
\begin{tabular}{lcc}
& C & R \\
T & 1, 4 & 1, 0 \\
M & 0, 0 & 3, 5 \\
B & 4, 4 & 0, 3
\end{tabular}
\end{table}

\onslide<2>{
$$
EU_1(0, 1/2, 1/2) =
\begin{cases}
2   & \text{ if } 2 \text{ chooses } C \\
1.5 & \text{ if } 2 \text{ chooses } R
\end{cases}
$$

}

## Example 

\begin{table}
\begin{tabular}{lcc}
& C & R \\
M & 0, 0 & 3, 5 \\
B & 4, 4 & 0, 3
\end{tabular}
\end{table}
